Solving the maximum independant set problem with einsum networks

### 2021-06-06 paper/paper.tex
fix all configurations
new masked configure finder

\begin{abstract}
	Solving the maximum independent set size problem by mapping the problem to an einsum network. 
    We obtain the maximum independent set size,
    the independence polynomial and the optimal configuration by using different tensor elements to a commutative semiring.
\end{abstract}

\section{Three approaches to computing independence polynomial}
The independence polynomial~\cite{Ferrin2014, Harvey2017} of a graph $G$ is defined as
\begin{equation}
I(G, x) = \sum_{k=1}^{\alpha(G)} a_k x^k,
\end{equation}
where $a_k$ is the number of independent sets of size $k$ in $G$, and $\alpha(G)$ is the maximum independent set size.
\blue{here, I preassumed a user knows tensor network, add more details later.}
If we map a graph into an einsum network, as shown in \Fig{fig:einsummapping}, by placing a rank one tensor of size $2$ at each vertex
\begin{equation}
    W(x)_{s_i} = \left(\begin{matrix}
        1 \\
        x
    \end{matrix}\right)_{s_i},
\end{equation}
and a rank two tensor of size $2 \times 2$ at each edge
\begin{equation}
    B_{s_i s_j} = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right)_{s_is_j},
\end{equation}
the contraction of this einsum network generates the independence polynomial %, we denote this contraction as \texttt{a,b,c,d,ab,ad,bc,bd,cd,ce->}.
\begin{equation}
    I(G, x) = \sum\limits_{s_1, s_2, \ldots, s_n = 0}^{1} \prod\limits_{i=1}^n W(x)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i s_j}.
\end{equation}
Here, the einsum enumerates all possible configurations and accumulates the product of tensor elements to the output.
The product over vertex tensors provides a factor $x^k$, where $k=\sum_i s_i$ is the set size.
While the product over edge tensors provides a factor $0$ if the configuration is not an independent set, otherwise $1$.
\begin{figure*}[t!]
    \centering
    \begin{subfigure}[t]{0.4\textwidth}
        \centering
        \centerline{\includegraphics[width=0.8\columnwidth,trim={0 0cm 0 0},clip]{../notebooks/fig1.pdf}}
        \caption{}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{0.4\textwidth}
        \centering
    \centerline{\includegraphics[width=0.8\columnwidth,trim={0 0cm 0 0},clip]{../notebooks/fig2.pdf}}
        \caption{}
    \end{subfigure}
    \caption{Mapping a graph to an einsum network.}\label{fig:einsummapping}
\end{figure*}
By mapping the problem to einsum, one can use some existing algorithms to decrease the complexity of the problem.
Especially those techiniques developed in recent quantum circuit simulations for finding the optimal exactly contraction order of a tensor network ~\cite{Gray2021,Pan2021}.
With these algorithms, one can easily find a good contraction order that reduces the complexity to a quantity related to the minimum treewidth of the graph.~\cite{Markov2008}
Such mapping also makes it possible to utilize faster BLAS functions.
Before contracting the tensor network and evaluating this polynomial numerically, let us first give up thinking $0$s and $1$s in tensors $W(x)$ and $B$ as regular computer numbers such as integers and floating point numbers.
Instead, we treat them as the additive identity and multiplicative identity of a commutative semiring.
A ring without additive inverse is called a semiring, and commutative semiring is a semiring that multiplication commutative.
To define a commutative semiring with addition algebra $\oplus$ and multiplication algebra $\odot$ on a set $R$, the following relation must hold for arbituary three elements $a, b, c \in R$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b + a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} * a = \mymathbb{0}
\end{align*}

@@

However, the result is very unreliable in practise due to the non-negligible round off error of floating point numbers.
Instead of choosing $X$ as a set of random real numbers, we let $x_i = r\omega^i$ and the above linear equation becomes
\begin{align}
\left(\begin{matrix}
1 & r\omega & r^2\omega^2 & \ldots & r^m\omega^m \\
1 & r\omega^2 & r^2\omega^4 & \ldots & r^m\omega^{2m} \\
\vdots & \vdots & \vdots &\ddots & \vdots \\
1 & r\omega^m & r^2\omega^{2m} & \ldots & r^m\omega^{m^2}
\end{matrix}\right)
\left(\begin{matrix}
a_0 \\ a_1 \\ \vdots \\ a_m
\end{matrix}\right)
= \left(\begin{matrix}
y_0 \\ y_1 \\ \vdots \\ y_m
\end{matrix}\right).
\end{align}

@@

\section{Computing maximum independent set size and its corresponding degeneracy and configurations}
Obtaining the maximum independent set size and its degeneracy is computational more efficient. Let $x=\infty$, then the independence polynomial becomes
\begin{equation}
I(G, \infty) = a_k \infty^{\alpha(G)},
\end{equation}
where the lower orders terms disappear automatically. We can define a new algebra as
\begin{align}
\begin{split}
    a_x\infty^x \oplus a_y\infty^y &= \begin{cases}
        (a_x + a_y)\infty^{\max(x,y)}, & x = y\\
        a_y\infty^{\max(x,y)}, & x < y\\
        a_x\infty^{\max(x,y)}, & x > y
    \end{cases}\\
    a_x\infty^x \odot a_y\infty^y &= a_x a_y\infty^{x+y}\\
    \mymathbb{0} &= 0\infty^{-\infty}\\
    \mymathbb{1} &= 1\infty^{0}
\end{split}
\end{align}
In the program, we only store the power $x$ and the corresponding factor $a_x$ that initialized to $1$.
This algebra is consistent with the one we derived in ~\cite{Liu2021} that uses the tropical tensor network solving spin glass ground states.
Here if we are only interested in obtaining $\alpha(G)$ and omit the factors, the algebra of $x$ corresponds to the tropical algebra~\cite{Maclagan2015,Moore2011}.

@@

>>>>

\begin{abstract}
	Solving the maximum independent set size problem by mapping the problem to an einsum network. 
    We obtain the maximum independent set size,
    the independence polynomial and the optimal configuration by generic programming tensor networks.
\end{abstract}

\section{Three approaches to compute independence polynomial}
The independence polynomial~\cite{Ferrin2014, Harvey2017} of a graph $G$ is defined as
\begin{equation}
I(G, x) = \sum_{k=1}^{\alpha(G)} a_k x^k,
\end{equation}
where $a_k$ is the number of independent sets of size $k$ in $G$, and $\alpha(G)$ is the maximum independent set size.
\blue{here, I preassumed a user knows tensor network, add more details later.}
If we map a graph into an einsum network, as shown in \Fig{fig:einsummapping}, by placing a rank one tensor of size $2$ at each vertex
\begin{equation}
    W(x)_{s_i} = \left(\begin{matrix}
        1 \\
        x
    \end{matrix}\right)_{s_i},
\end{equation}
and a rank two tensor of size $2 \times 2$ at each edge
\begin{equation}
    B_{s_i s_j} = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right)_{s_is_j},
\end{equation}
the contraction of this einsum network gives the independence polynomial %, we denote this contraction as \texttt{a,b,c,d,ab,ad,bc,bd,cd,ce->}.
\begin{equation}
    I(G, x) = \sum\limits_{s_1, s_2, \ldots, s_n = 0}^{1} \prod\limits_{i=1}^n W(x)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i s_j}.
\end{equation}
Here, the einsum enumerates all possible configurations and accumulates the product of tensor elements to the output.
The product over vertex tensors provides a factor $x^k$, where $k=\sum_i s_i$ is the set size.
While the product over edge tensors provides a factor $0$ if the configuration is not an independent set, otherwise $1$.
\begin{figure*}[t!]
    \centering
    \begin{subfigure}[t]{0.4\textwidth}
        \centering
        \centerline{\includegraphics[width=0.8\columnwidth,trim={0 0cm 0 0},clip]{../notebooks/fig1.pdf}}
        \caption{}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{0.4\textwidth}
        \centering
    \centerline{\includegraphics[width=0.8\columnwidth,trim={0 0cm 0 0},clip]{../notebooks/fig2.pdf}}
        \caption{}
    \end{subfigure}
    \caption{Mapping a graph to an einsum network.}\label{fig:einsummapping}
\end{figure*}
By mapping the problem to einsum, one can use some existing algorithms to decrease the complexity of the problem.
Especially those techiniques developed in recent quantum circuit simulations for finding the optimal exactly contraction order of a tensor network ~\cite{Gray2021,Pan2021}.
With these algorithms, one can easily find a good contraction order that reduces the complexity to a quantity related to the minimum treewidth of the graph.~\cite{Markov2008}
Such mapping also makes it possible to utilize faster BLAS functions.
Before contracting the tensor network and evaluating this polynomial numerically, let us first give up thinking $0$s and $1$s in tensors $W(x)$ and $B$ as regular computer numbers such as integers and floating point numbers.
Instead, we treat them as the additive identity and multiplicative identity of a commutative semiring.
A ring without additive inverse is called a semiring, and commutative semiring is a semiring that multiplication commutative.
To define a commutative semiring with addition algebra $\oplus$ and multiplication algebra $\odot$ on a set $R$, the following relation must hold for arbituary three elements $a, b, c \in R$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b + a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} * a = \mymathbb{0}
\end{align*}

@@

However, the result is very unreliable in practise due to the non-negligible round off error of floating point operations.
Instead of choosing $X$ as a set of random real numbers, we let $x_i = r\omega^i$ and the above linear equation becomes
\begin{align}
\left(\begin{matrix}
1 & r\omega & r^2\omega^2 & \ldots & r^m\omega^m \\
1 & r\omega^2 & r^2\omega^4 & \ldots & r^m\omega^{2m} \\
\vdots & \vdots & \vdots &\ddots & \vdots \\
1 & r\omega^m & r^2\omega^{2m} & \ldots & r^m\omega^{m^2}
\end{matrix}\right)
\left(\begin{matrix}
a_0 \\ a_1 \\ \vdots \\ a_m
\end{matrix}\right)
= \left(\begin{matrix}
y_0 \\ y_1 \\ \vdots \\ y_m
\end{matrix}\right).
\end{align}

@@

\section{Computing maximum independent set size and its corresponding degeneracy and configurations}
Obtaining the maximum independent set size and its degeneracy is computational more efficient. Let $x=\infty$, then the independence polynomial becomes
\begin{equation}
I(G, \infty) = a_k \infty^{\alpha(G)},
\end{equation}
where the lower orders terms disappear automatically. We can define a new algebra as
\begin{align}
\begin{split}
    a_x\infty^x \oplus a_y\infty^y &= \begin{cases}
        (a_x + a_y)\infty^{\max(x,y)}, & x = y\\
        a_y\infty^{\max(x,y)}, & x < y\\
        a_x\infty^{\max(x,y)}, & x > y
    \end{cases}\\
    a_x\infty^x \odot a_y\infty^y &= a_x a_y\infty^{x+y}\\
    \mymathbb{0} &= 0\infty^{-\infty}\\
    \mymathbb{1} &= 1\infty^{0}
\end{split}
\end{align}
In the program, we only store the power $x$ and the corresponding factor $a_x$ that initialized to $1$.
This algebra is consistent with the one we derived in ~\cite{Liu2021} that uses the tropical tensor network solving spin glass ground states.
Here if we are only interested in obtaining $\alpha(G)$ and omit the factors, the algebra of $x$ corresponds to the max-plus tropical algebra~\cite{Maclagan2015,Moore2011}.

@@

\subsection{bounding the enumeration space}
If one tries to implement the above algebra for enumerating configurations, he will find the space overhead is not propotional to the degeneracy of the optimal configuration.
It stores more than nessesary intermedite configurations. To speed up calculation, we use the result of much faster tropical number for bounding.
We first compute the optimal maximum independent set size $\alpha(G)$ with tropical numbers and cache all intermediate tensors in the contraction route.
Then we use boolean masks to mark relevent items for each cached tensor, i.e. using a boolean true to mark a tensor element with nonzero gradients and boolean false otherwise.
Finally, we perform masked matrix multiplication using the new element type with the above algebra for obtaining all configurations.
To compute the masks, we ``back propagate'' the masks through contraction route.
Consider a tropical matrix multiplication $C = A B$, we have the following inequality
\begin{align}
    A_{ij} \odot B_{jk} &\leq C_{ik}
\end{align}

Using the tropical algebra, we have
\begin{align}
    A_{ij} &\leq (\max_{k} (C_{ik}^{-1} \odot B_{jk}))^{-1}
\end{align}
where the tropical multiplicative inverse is defined as the additive inverse of the regular algebra. The equality holds if and only if $A_{ij}$ can contribute to $C$ (i.e. has nonzero gradient).
Let the mask for $C$ being $\overline C$, the backward rule for the mask reads
\begin{align}
\overline{A}_{ij} = \delta(A_{ij}, ((C^{\circ-1} \circ \overline C )B^T)_{ij}^{\circ -1}),
\end{align}
where ${}^{\circ -1}$ is the Hadamard inverse, $\circ$ is the Hadamard multiplication, boolean false is treated as tropical zero and boolean true is treated as tropical one.
This backward rule utilizes the properties that tropical algebra has multiplicative inverse for a nonzero element, while the ordering is well defined too.
This rule defined on matrix multiplication can be easily generalized to the einsum by replacing the matrix multiplication between $C^{\circ-1} \circ \overline C$ and $B^T$ by an einsum.
